A Collection of Thoughts on the Keccak Construction

### 2018-04-12 s2018-math490/thesis/thesis.tex
Changes to thesis
Signed-off-by: Alexander Scheel <alexander.m.scheel@gmail.com>

    While there have been many applications of mathematics, few are as
demanding and shrouded in as much secrecy as cryptography. Cryptography exists
because of the fundamental need of civilizations, governments, and individuals
to keep secrets secure from devoted adversaries. While modern cryptography
combines the disciplines of mathematics, computer science, and computer
engineering, prior to the turn of the 20th century, cryptography lacked much
of its modern rigor.

    Prior to the 20th century, cryptography was split into three major
branches: ciphers, codes, and stenography.

@@

@@

    SHA-3 consists of two parts: a core permutation function, KECCAK-$f$, and
a domain extender, the KECCAK sponge function.

@@

Define $\Sigma = \{0, 1\}$ to be the alphabet, and for
$w \in \{1, 2, 4, 8, 16, 32, 64\}$, $S_{w} = \Sigma^{25w}$, to be the set of
all binary strings of length $25w$.

@@

Lastly, we have computed the order of the permutation $\theta$ for all $w$.
We reproduce them here without proof; they were found by direct search, but
only verified exhaustively for $w = 1$. In general, the order is given by
the expression $3 \times w$.

@@

\subsection{Evaluation of the Orders of Composition of Permutations}

>>>>

    While there are many applications of pure mathematics, few are as
demanding and shrouded in secrecy as cryptography. Cryptography exists
because of the fundamental need of civilizations, governments, and individuals
to keep secrets secure from devoted adversaries. While modern cryptography
combines the disciplines of mathematics, computer science, and computer
engineering, prior to the turn of the 20th century, cryptography lacked much
of its modern rigor.

    Prior to the late 19th century, cryptography was split into three major
branches: ciphers, codes, and stenography.

@@

This section contains a collection terminology useful for discussing SHA-3
and Boolean Satisfiability.

We define $\Sigma = \{0, 1\}$ to be the alphabet.

We define the set $W = \{1, 2, 4, 8, 16, 32, 64\}$ to be the powers of two
typically used in constructing the Keccak widths; $w=64$ is standardized for
use in SHA-3, though any power of two can be used.

For a given $w \in W$, we define $S_{w} = \Sigma^{25w}$, to be the set of
all binary strings of length $25w$; these represent the possible states (which
are binary strings that map onto an indexible array $A$ described later) and each
round permutation maps $S_{w} \mapsto S_{w}$.

A binary string $s \in S_{w}$ can be indexed directly (in a zero-indexed
manner, i.e., the starting bit of $s$ is denoted $s[0]$), or via a 3
dimensional structure of size 5 by 5 by $w$. This is typically done in
conjunction with an uppercase letter, $A = s$, and then
$A[x, y, z] = s[w(5y + x) + z]$. This is in accordance with
FIPS 202 \cite{NIST202}.

@@

    SHA-3 consists of two parts: a core permutation function, KECCAK-$f$, and
a domain extender, the KECCAK sponge function. As discussed by the Keccak
authors, additional security is given by choosing the permutation functions to
be bijective, though they need not be (the domain need only be $S_{w}$).

@@

@@

Lastly, we have computed the order of the permutation $\theta$ for all $w$.
We reproduce them here without proof; they were found by randomized search, and
verified with SAT for $w = 1, 2, 4$ and $8$. In general, the order is given by
the expression $3 \times w$.

@@

\subsection{Evaluation of the Orders of Composition of Permutations} \label{sec:p:composition}

In this section, we discuss how the above five permutation functions compose,
and the orders of the resulting compositions. We limit our discussion to $w=1$
for the interests of exhaustive search: $2^{25}$ is possible on commodity
hardware, $2^{50}$ would require additional resources. In general, we find that
these permutations interact non-trivially, resulting in cycles of mixed size.
We reproduce these results in the table below:

\begin{tabular}{c c c c} \label{tab:p:c:1}
    Function & Order & Number of Cycle & List of Cycles \\
\end{tabular}
